Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging

### 2012-06-04 HRUIonGPU/jrnl_article.tex
done some writing on tuffc article

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
\IEEEPARstart{C}{apon} adaptive beamforming applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.  In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied on typical cardiac data sets. To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows. Chen et al. \cite{Chen2011a, Chen2011} have previously implemented the Capon beamformer on a GPU, however their implementation was based on real data, where the Capon beamformer assumes complex. Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced compared with use of complex data. It is therefore crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows.

@@

\section{Background}
\subsection{Capon Beamforming}
When sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming. The conventional form of beamforming is delay-and-sum (DAS), where the signal recorded on each element, $x_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
In conventional beamforming, the weight $w_m \in \mathbb{R}$ is usually applied to trade resolution for a better signal-to-noise ratio (SNR). Hence, by increasing the main lobe width of the system while gaining lower side lobe levels.

The basic idea of an adaptive beamformer is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria in such a way that we minimize interference impinging from other directions than the point of focus. For a complex data vector $\vec{x}[n] \in \mathbb{C}^M$, we have for an adaptive beamformer a data-dependent set of weights, $\vec{w}(\vec{x}[n])$, that when applied on $\vec{x}[n]$ is the best approximation to achieving the selected criteria. With the Capon beamformer, adaptive weights are found by solving the following minimization problem \cite{Capon1969}:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{eq:constraint}
\end{align}
where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a}$ is the far-field steering direction.
Thus, we minimize the output power, while maintaining unit gain in the look direction. The constraint in Eq. \ref{eq:constraint} is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$. The solution to the minimization problem in Eq. \ref{eq:minz} and \ref{eq:constraint} is
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
\end{align}
where $\vec{a} = \vec{1}$ when $\vec{x}$ is pre-delayed to broadside.

In practice, the covariance matrix has to be approximated by the sample covariance matrix, $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$. In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over all available samples to improve on the estimate \cite{Krima}. For an active, pulsed system like ultrasound imaging, the available data from which $\mat{R}$ can be estimated is quite limited. The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned. Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favours weights that in the end cancel out the signal \cite{Synnevag2009a}. In this way, the beamformer output can be zero at broadside, even when there is signal present and the distortion-less criterion is applied. 
In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N$ time samples \cite{Synnevag2007, Synnevag2007a}.
\begin{align}
\mat{\hat{R}} = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where $K = M-L+1$, $N = 2N_{avg} + 1\sim \tau/T_s$ ($\tau$ is the pulse length and $T_s$ is the sampling period) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, 
\begin{align}
\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}.
\end{align}
A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\hat{R}}$, 
\begin{align}
\epsilon = d*tr\{\mat{\hat{R}}\},
\end{align}
has previously shown good results in the literature \cite{Synnevag2007}.

@@

\subsection{GPU Compute Model}
In this section we briefly discuss the principle behind GPU computing. CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}. However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper applies for other massive parallel implementation environments like e.g. OpenCL as well.

The CUDA compute model is based on execution of a kernel-function across a grid of compute threads. Each position in the grid holds a block of threads.
%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
From a CUDA perspective the GPU consist of one or more stream multi processors (SM's), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. A group of blocks is scheduled to each SM until all blocks have been processed. More important, the SM has a limited amount of near-core memory and registers available. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency. This pose a real challenge as the target problem then has to map to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.

@@

\section{Parallel Capon}\label{sec:meth}
As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity. However, the calculation of Capon weights is well suited for a GPU implementation, since the calculations are both independent across pixels, and partially intra-pixel. On a top level it is trivial to recognise that one grid cell of threads should process one pixel in the final image. Continuing, on a pixel-level, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes. For each data vector $\vec{x}$ in the image:
\begin{enumerate}
\item Build the sample covariance matrix $\mat{\hat{R}}$
\item Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
\item Calculate $\vec{w} = \vec{b}/|\vec{b}|$ and the amplitude $\vec{w}^H\vec{x}$
\end{enumerate}
The computational complexity of these three steps, with $L = M/2$, is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively. For larger $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by finding novel solutions to these two steps.

\subsection{Building Multiple Sample Covariance Matrices}
The straight forward sum in Eq. \ref{eq:R} has complexity $O(L^2KN) = O(L^3)$ if $L$ is large. Note that N typically is small compared to L and M. For smaller $L$ the construction of $\mat{\hat{R}}$ is therefore dominated by $M$. The quadratic complexity for large $L$ can be improved by exploiting the overlap from element to element on the diagonals of $\mat{\hat{R}}$ (ref figure). This overlap is present both across subarrays and time.

@@

\subsection{Beamspace Processing}
Nilsen et. al \cite{Nilsen2009} has proposed to apply beamspace processing to further reduce the required computations. With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.  The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space. We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$. For a focused system like a cardiac ultrasound scanner, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy. Nilsen et. al concluded that as little as three beams around broadside where adequate to produce results comparable to applying Capon beamforming in element space. The beamspace version of Capon is derived by replacing the covariance matrix in Eq. \ref{eq:w} with the beamspace covariance matrix $\mat{R}_{BS} = \vec{x}_{BS}\vec{x}_{BS}^H$.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
\bibliography{mybib} % link to mendeley organized bibtex-file
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}

>>>>

%\IEEEPARstart{I}{ntroduction} to Capon adaptive beamforming, ultrasound imaging and GPU/CUDA.
\IEEEPARstart{C}{apon} adaptive beamforming applied in a medical ultrasound setting has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. Although the method introduce some interesting trade-offs in addition to just improved resolution, like maintaining image resolution combined with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}, widespread use has been limited by the tremendous computational complexity associated with the method \cite{So2011}.  In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied on a typical cardiac data set. To accomplish this task we make use of the massive parallel processing power found in modern graphics processing units (GPU's), combined with novel methods to reduce the computational complexity as the problem size grows. Earlier attempts on real time Capon beamforming have been focused around radar and passive sonar using intricate systolic array processors \cite{McWhirter1989, Moonen1993, Sinha2002}.  A systolic array processor is custom made hardware, and even if the computational power might be higher for selected problems compared with a GPU, the cost related to development and flexibility is much higher. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011} have previously implemented the Capon beamformer on a GPU, however their implementation was based on real data, where the Capon beamformer assumes complex. Using real data will results in symmetric windows, and even if the number of samples they managed to process per second was high, both the number of floating point operations per sample and the imaging region is reduced compared with use of complex data. It is therefore crucial to use complex data within the Capon beamformer framework, both for reducing the amount of data per image, and to allow for asymmetric windows. All results in this paper is reported with use of critically sampled complex data, both laterally and in range.

@@

\section{Background}
\subsection{Capon Beamforming}
When sensing the world around us with an array of elements, we perform a spatio-temporal filtering operation known as beamforming. The conventional form of beamforming is delay-and-sum (DAS), where the signal recorded on each element, $x_m[n]$, is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n].
\end{align}
In conventional beamforming, the weighting $\vec{w} \in \mathbb{R}$ is usually applied to trade resolution for a better signal-to-noise ratio (SNR). Hence, by increasing the main lobe width of the system while gaining lower side lobe levels.

The basic idea of an adaptive beamformer is to calculate the weights on-the-fly, based on the data at hand, and on a selected optimization criteria in such a way that we minimize interference impinging from other directions than the point of focus. For a complex data vector $\vec{x}[n] \in \mathbb{C}^M$, we have for an adaptive beamformer a data-dependent set of weights, $\vec{w}(\vec{x}[n])$, that when applied on $\vec{x}[n]$ is the best approximation to achieving the selected criteria. With the Capon beamformer, adaptive weights are found by solving the following minimization problem \cite{Capon1969}:
\begin{align}
&\min_{\vec{w}} E\{|z[n]|^2\} = \min_{\vec{w}}\vec{w}^H\mat{R}\vec{w}\label{eq:minz}\\
&\text{subjected to } \vec{w}^H\vec{a} = 1,\label{eq:constraint}
\end{align}
where $\mat{R} = E\{\vec{x}[n]\vec{x}[n]^H\}$ is the spatial covariance matrix, and $\vec{a} = \left[ \begin{array}{cccc} e^{-jk_{\tilde{x}}\tilde{x}_0}& e^{-jk_{\tilde{x}}\tilde{x}_1}& ...& e^{-jk_{\tilde{x}}\tilde{x}_{M-1}}\end{array}\right] ^T$ is a far-field steering vector towards angle $\theta = \arcsin(\frac{\lambda k_{\tilde{x}}}{2\pi})$.
Thus, we minimize the output power, while maintaining unit gain in the look direction. The constraint in (\ref{eq:constraint}) is there to avoid the trivial solution $\vec{w} = 0$, and a distorted respons in the direction of $\vec{a}$. The solution to the minimization problem in (\ref{eq:minz}) and (\ref{eq:constraint}) is
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\mat{R}[n]^{-1}\vec{a}}{\vec{a}^H\mat{R}[n]^{-1}\vec{a}},
\end{align}
where $\vec{a} = [1_0 1_1... 1_{M-1}]^T = \vec{1}$ when $\vec{x}$ is pre-delayed to broadside.

In practice, the covariance matrix has to be approximated by the sample covariance matrix, $\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H$. In passive systems, assuming stationary statistics, the sample covariance matrix is typically averaged over all available samples to improve on the estimate \cite{Krima}. For an active, pulsed system like ultrasound imaging, the available data from which $\mat{R}$ can be estimated is quite limited. The resulting sample covariance matrix might therefore be both inaccurate, and poorly conditioned. Another challenge with applying Capon to active systems is the effect of signal cancellation, where high correlation between signal and interference favours weights that in the end cancel out the signal \cite{Synnevag2009a}. In this way, the beamformer output can be zero at broadside, even when there is signal present and the distortion-less criterion is applied. 
In order to get a well conditioned $\mat{\hat{R}}$, avoid signal cancellation and to get DAS-like speckle, $\mat{\hat{R}}$ has to be averaged over $L\le M/2$ long subarrays and $N$ time samples \cite{Synnevag2007, Synnevag2007a}.
\begin{align}
\mat{\hat{R}}[n] = \frac{1}{NK}\sum_{n'=n-N_{avg}}^{n+N_{avg}} \sum_{l=0}^{K-1} \vec{x}_l[n']\vec{x}_l[n']^H,\label{eq:R}
\end{align}
where $K = M-L+1$, $N = 2N_{avg} + 1\sim \tau/T_s$ ($\tau$ is the pulse length and $T_s$ is the sampling period) and $\vec{x}_l$ is the $l^\text{th}$ subarray $[x_l[n], \dotso x_{l+L}[n]]$.
Finally, $\mat{\hat{R}}$ is loaded with diagonal factor $\epsilon$, 
\begin{align}
\mat{\hat{R}} = \mat{\hat{R}} + \epsilon\mat{I}.
\end{align}
A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\hat{R}}$, 
\begin{align}
\epsilon &= d*tr\{\mat{\hat{R}}\}, \\
d &= 1/L
\end{align}
introduced by Synnev\aa{}g et al. \cite{Synnevag2007} has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Nilsen2009, Wang2009, Mehdizadeh2012}.

@@

\subsection{GPU Compute Model}
In this section we briefly discuss the principle behind GPU computing, laying the basis for further discussion on parallel implementation strategies. CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}. However, even though CUDA has been used to implement the Capon beamformer, the challenges and proposed methods described in this paper applies for other massive parallel implementation environments as well, like e.g. OpenCL.

The CUDA compute model is based on execution of a kernel-function across a grid of compute threads. Each position in the grid holds a block of threads.
%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
From a CUDA perspective the GPU consist of one or more stream multi processors (SM's), where each SM is capable of 32 or more, depending on the architecture, concurrent multiplications and/or additions. A group of blocks is scheduled to each SM until all blocks have been processed. More important, the SM has a limited amount of near-core memory and registers available. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency. This pose a real challenge as the target problem has to map to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration.

@@

\section{Parallel Capon}\label{sec:meth}
As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity. However, the calculation of Capon weights is well suited for a GPU implementation, since the calculations are both independent across pixels, and partially intra-pixel. Based on memory dependencies it is trivial to recognise that one block of threads should process at least one pixel in the final image. Continuing, on a pixel-level, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes. That is, for each data vector $\vec{x}$ in the image:
\begin{enumerate}
\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
\item Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
\item Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
\end{enumerate}
The computational complexity of these three steps, with $L = M/2$, is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively. For larger $L$'s, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.

\subsection{Building Multiple Sample Covariance Matrices}
The straight forward sum in (\ref{eq:R}) has complexity $O(L^2KN) = O(L^3)$ if $L$ is a small fraction of $M$. This is actually the same as for inverting a $L \times L$ matrix. However, little attention has been devoted to this step in the literature until now. Note that for a typical array configuration, N is usually small compared to L and M. Hence, for smaller $L$ the construction of $\mat{\hat{R}}$ is dominated by $M$, since $K = M - L + 1$.

The quadratic complexity with $L$ and linear complexity with $N$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time. The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j > 0$ by subtracting $x_{i-1}x_{j-1}^*$ and adding $x_{i+K-1}x_{j+K-1}^*$. The same is true in time where $\mat{\hat{R}}[n]$ can be calculated from $\mat{\hat{R}}[n-1]$ by subtracting $\mat{\hat{R}}[n-N_{avg}-1]$ and adding $\mat{\hat{R}}[n+N_{avg}]$. Both sliding across subarrays and in time can be realized simultaneously if we hold $\vec{x}\vec{x}^H[n]$ averaged over $N$ time samples in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (figure). However, since $N$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$.

Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated. However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU. This also means that we do not relay on a symmetric solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}$, even if that could have saved some instructions. The memory layout for the upper triangle only is also cumbersome to work with.

For subarray averaging, the first row and column of $\mat{\hat{R}}$ has to be fully calculated, and for time averaging $\mat{\hat{R}}[0]$ has to be fully calculated.

Since $L$ usually is larger than $N$, we have

@@

\subsection{Beamspace Processing}
Nilsen et. al \cite{Nilsen2009} has proposed to apply beamspace processing to further reduce the required computations. With beamspace we mean transforming the channel data from element space to a fan of beams covering the imaging sector, $\vec{x}_{BS} = \mat{B}x$.  The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space. We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$. For a focused system like a cardiac ultrasound scanner, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy. Nilsen et. al concluded that as little as three beams around broadside where adequate to produce results comparable to applying Capon beamforming in element space. The beamspace version of Capon is derived by replacing the covariance matrix in (\ref{eq:w}) with the beamspace covariance matrix $\mat{R}_{BS} = \vec{x}_{BS}\vec{x}_{BS}^H$, where $\mat{B} \in \mathbb{C}^{N_b,M}$, and $N_b$ is the number of selected beams. Note that when subarray averaging is applied, the transformation must be done per subarray, yielding subarrays that no longer overlap in space. Hence, the amount of data is change from $M$ to $KN_b$ per sample.

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
\bibliography{mybib} % link to mendeley organized bibtex-file
%\bibliography{IEEEabrv,mybib} 
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
%\begin{thebibliography}{1}
